---
title: "3D-Contour plots"
author: "Yajie Guo"
date: "20/01/2025"
output: html_document
---
output:
  html_document:
    extra_dependencies:
      - amsmath
      - someotherpackage
      - extarrows
      - color
      - amssymb
      - cases
---

#The loglikelihood of $\theta_1,\theta_2$ only related to total birth number of protein and the birth time.

$$
L(\theta_1,\theta_2)=ln(\theta_2)\sum_{i=1}^k e_i-\theta_1\sum_{i=1}^k e_it_i
 + \frac{\theta_2x_1(0)}{\theta_1}(exp(-\theta_1T)-1)
 $$


```{r}

library('dplyr')


ssa.f<-function(Mt,theta,S0){
  
  Sini<-S0
  
  names(Sini)<-c('S1','S2')
  
  S.can<-S0
  
  t<-0
  
  Reaction<-matrix(c(-1,0,0,+1,0,-1),byrow=TRUE,ncol=2)
  
  while(t[length(t)] <= Mt){
    
    if ( S.can[1] != 0  |   S.can[2] != 0 ){
      
      theta.can<-c(theta[1]*S.can[1], theta[2]*S.can[1], theta[3]*S.can[2])
      
      lambda<-sum(theta.can, na.rm = T) 
      
      wait<-rexp(1,rate=lambda)
      
      
      t[length(t)+1] <-t[length(t)]+wait
      
      if ( S.can[1] > 0  ){
        
        
        prob<-theta.can/lambda   
        
        R.can<-sample(size =1,x=c(1,2,3), prob = prob)
        
        
        
        if(R.can==1){
          S.can<-S.can+Reaction[1,] } else{
            
            if(R.can==3){
              S.can<-S.can+Reaction[3,]
            }else{
              
              S.can<-S.can+Reaction[2,]
            }
          }
        
      }else{
        
        if(S.can[2] > 0){
          S.can<- S.can+Reaction[3,] }else{ return( cbind.data.frame(t,Sini) )  } 
        
      }
      
    }else{ return(cbind.data.frame(t,Sini) )}
    
    Sini<-rbind.data.frame(Sini,S.can)
    
  }
  
  
  names(Sini)<-paste0('Species',c('mRNA','Protein'))
  cbind.data.frame(jump.times=t,Sini)
  
}

#original function,ssa
ori.f<-function(Mt,theta,S0){
  
  Sini<-S0
  
  
  S.can<-S0
  
  t<-0
  
  Reaction<-matrix(c(-1,0,0,+1,0,-1),byrow=TRUE,ncol=2)
  
  
  
  while(t[length(t)] <  Mt){
    
    
    if ( S.can[1] != 0  |   S.can[2] != 0 ){
      
      
      theta.can<-c(theta[1]*S.can[1], theta[2]*S.can[1], theta[3]*S.can[2])
      
      lambda<-sum(theta.can, na.rm = T) 
      
      wait<-rexp(1,rate=lambda)
      
      
      
      t[length(t)+1] <-t[length(t)]+wait
      
      
      if(t[length(t)]<= Mt){
        
        if ( S.can[1] > 0  ){
          
          
          prob<-theta.can/lambda   
          
          R.can<-sample(size =1,x=c(1,2,3), prob = prob)
          
          
          
          if(R.can==1){
            S.can<-S.can+Reaction[1,] 
            
            
          }else{
            
            if(R.can==3){
              #protein death process
              
              
              
              S.can<-S.can+Reaction[3,]
              
              
            }else{
              
              S.can<-S.can+Reaction[2,] #protein birth process
            }
          }
          
        }else{
          
          if(S.can[2] > 0){
            S.can<- S.can+Reaction[3,]  #protein death process
            
          }
          
        }
        
      }else{
        t[length(t)]=Mt
        
      }   
      
    }
    
    
    Sini<-rbind.data.frame(Sini,S.can)
    
  }
  
  total.d<- cbind.data.frame(t,Sini)
  
  colnames(total.d)<-c("t","S1.ori","S2.ori")
  
  return(total.d)
  
}



#Extract the Ei(event.type up or down) and Ti(integral of integral) from original Protein process(S2) 

OriS2.ei.ti<-function(original.out){
  
  S2.ori<-original.out[,3]
  
  Ei<-rep(NA,l=length(S2.ori))
  Ti<-original.out[,1]
  
  ori.I2<-0
  
  
  for(i in 2:(length(S2.ori)-1)){
    
    if(S2.ori[i] == (S2.ori[i-1]+1)){
      Ei[i]<-1 #birth_of_protein
      
    }else{
      if(S2.ori[i] == (S2.ori[i-1]-1)){
        Ei[i]<-0  #death_of_protein
        
      }else{
        Ei[i]<-NA 
        Ti[i]<-NA
      }}
  }
  
  
  
  #the integral of protein
  
  index<-which(!is.na(Ti))
  
  
  for(i in 2:length(index)){
    
    ori.I2<-ori.I2+(Ti[index[i]]-Ti[index[i-1]])*S2.ori[index[i-1]]
  }
  
  S2.data.na<-cbind(Ti,S2.ori,Ei)
  
  S2.data<-S2.data.na[index,]
  
  list(total.data=S2.data[,1:2],Event.type=S2.data[,3],Integral.protein=ori.I2)
}



```


#Generate one dataset of CTMC model for the original mRNA and protein system

```{r}


N=1000
#Boundary.parameter<-c(zapsmall(0+(1e-5), digits = 6),zapsmall(1-(1e-5), digits = 6))

theta<-c(0.11,0.001*N,0.03) 

S0<-c(N,0)

Mt=50


result<-na.omit(ssa.f(Mt,theta,S0)) 




result[length(result[,1]),]<-c(Mt,result[length(result[,1])-1,2],result[length(result[,1])-1,3])



S.pro<- result %>% select(c('SpeciesProtein'))

ei<-rep(NA,l=length(S.pro[,1]))

for(j in 2:(length(S.pro[,1])-1)){
  ei[j]<-S.pro[j,1]-S.pro[j-1,1]
}

sum.ei<- length(which(ei==1))
index.birth<- which(ei==1)

birth.ti<-result$jump.times[index.birth]  #birth time, jumps times

```

#Extract the data of sum_ei(total birth number) and birth time(t_i) in likelihood function



```{r}

##Parameter Estimator

#input the (theta_1, log(theta_2) and log(theta_3))

Neg.loglike.fun<-function(theta,ei,Ti,S0.1,I.2){
  
  theta1<-theta[1]
  
  # ln.theta2<-log(theta[2])
  
  ln.theta2<-theta[2]
  
  theta2<-exp(theta[2])
  
  # ln.theta3<-log(theta[3])
  
  ln.theta3<-theta[3]
  theta3<-exp(theta[3])
  
  t.max=Mt
  OUT<- -( sum(ei)*ln.theta2- theta1*sum(ei*Ti)+ sum(1-ei)*ln.theta3+ ((theta2*S0.1)/theta1)*(exp(-theta1*t.max)-1)-theta3*I.2)
  
  return(OUT)
}



MLE.es<-function(single.out,S0){
  
  Ti<-single.out$total.data[,1][-c(1,length(single.out$total.data[,1]))]
  
  ei<-single.out$Event.type[-c(1,length(single.out$Event.type))]
  
  S0.1<-S0[1]
  
  I.2<-single.out$Integral.protein
  
  #hat.mle<-optim(par=c(0.01,0.01,0.01),fn=Neg.loglike.fun,Ti=Ti,ei=ei,S0.1=S0.1,I.2=I.2,lower = Boundary.parameter[1], upper =Boundary.parameter[2],method="L-BFGS-B")$par
  
  
  hat.ln.mle<-optim(par=c(0.01,0.01,0.01),fn=Neg.loglike.fun,Ti=Ti,ei=ei,S0.1=S0.1,I.2=I.2)$par
  
  hat.mle<-c(hat.ln.mle[1],exp(hat.ln.mle[2:3]))                 
  return(hat.mle)
}
```



```{r}
#write.csv(result, file = "myProtein-one-trajectory.csv", row.names = FALSE)
data.llk.jumptime<-OriS2.ei.ti(result)

#saveRDS(data.llk.jumptime, "myProteinLoglikedata.csv")

#mydata2 <- readRDS("myProteinLoglikedata.csv")

(hat.theta12<-MLE.es(single.out = data.llk.jumptime,S0=S0)[1:2])

```




#Start here!



#Run the MLE from the given one trajectory





```{r}
#given data of sum.ei and birth time ti
loglikelihood.full.data<-function(par1,par2){
  
  the1<- par1
  
  the2<- par2
  ln2<-log(the2)
  
  x0<-S0[1]
  
  out<- sum.ei*ln2-the1*sum(birth.ti)+(the2*x0*(exp(-the1*Mt)-1))/the1
  #out<100*out
  return(out)
}

```

#Start from here: 3D plot of loglikelihood function

```{r}

par_true <- c(0.11, 1)


# Zoom range around (0.11, 1)
x_zoom <- seq(par_true[1] - 0.02, par_true[1] + 0.02, length.out = 200)
y_zoom <- seq(par_true[2] - 0.1, par_true[2] + 0.1, length.out = 200)


# Recompute z for the zoomed grid
z_zoom <- outer(x_zoom, y_zoom, loglikelihood.full.data)

```






```{r}


pdf(file="myContour-with-TURE-Theta.pdf", width = 8, height = 6)
contour(
  x_zoom, 
  y_zoom, 
  z_zoom,
  xlab  = expression(theta[1]),    # θ₁
 ylab = expression(theta[2]),    # θ₂
  main = "Log-Likelihood Contour Plot",
 nlevels=80
)

# Mark the true parameter in 2D
points(
  x = par_true[1], 
  y = par_true[2], 
  pch = 19, 
  col = "red", 
  cex = 1
)
points(
  x =hat.theta12[1], 
  y = hat.theta12[2], 
  pch = 19, 
  col = "blue", 
  cex = 1
)


text(
  x     = par_true[1],
   y = par_true[2], 
  labels = bquote(
  theta^{"true"} == "(" ~ .(round(par_true[1], 3)) ~ "," ~ .(round(par_true[2], 3)) ~ ")"
),
  pos   = 4,
  col   = "red"
)

text(
  x     = hat.theta12[1],
   y = hat.theta12[2], 
  labels = bquote(
  theta^{"mle"} == "(" ~ .(round(hat.theta12[1], 3)) ~ "," ~ .(round(hat.theta12[2], 3)) ~ ")"
),
  pos   = 4,
  col   = "blue"
)


dev.off()  # this line finalizes and closes the PDF device
```




#3D loglikelihood plots

```{r}
library(plot3D)

library(plot3Drgl)

#x <- seq(0.08, 0.14, length = 1000) #theta1 0.11 true 
#y <- seq(0.8,1.2, length = 1000) #theta2  1 true
#z <- outer(x, y, loglikelihood.full.data)

t.l<-loglikelihood.full.data(par1=0.11,par2=1)

# xlab=expression(theta[2]),ylab=expression(theta[1]) 

# plot the 3D surface
persp3d(x_zoom, y_zoom, z_zoom,col = "lightblue", shade = 0,border = NA,theta = 90,xlab='theta1',ylab='theta2',zlab='loglikelihood',box = FALSE)


points3d(
  x =hat.theta12[1], 
  y = hat.theta12[2], 
  z = loglikelihood.full.data(par1=hat.theta12[1],par2=hat.theta12[2]),
  size = 10, 
  col = "blue", 
  cex = 1
)

points3d(
  x = par_true[1],
  y = par_true[2],
  z = t.l,
  color = "red",
  size = 10
)



# -- (2a) Add Contour Lines on the Floor (optional) --
#z_floor <- min(z, na.rm = TRUE)
#contour_levels <- pretty(range(z, na.rm = TRUE), n = 10)
#contours <- contourLines(x, y, z, levels = contour_levels)
#for (i in seq_along(contours)) {
#  xs <- contours[[i]]$x
#  ys <- contours[[i]]$y
#  lines3d(xs, ys, rep(z_floor, length(xs)), color = "black", lwd = 2)
#}
# Recompute z for the zoomed grid

z_floor <- min(z_zoom, na.rm = TRUE)

# Create more contour levels (e.g., 30)
range_z_zoom <- range(z_zoom, na.rm = TRUE)
contour_levels_zoom <- seq(range_z_zoom[1], range_z_zoom[2], length.out = 60)

# Get contour lines at these finer levels
contours_zoom <- contourLines(x_zoom, y_zoom, z_zoom, levels = contour_levels_zoom)

# Draw the floor contours in rgl
for (i in seq_along(contours_zoom)) {
  xs <- contours_zoom[[i]]$x
  ys <- contours_zoom[[i]]$y
  lines3d(xs, ys, rep(z_floor, length(xs)), color = "black", lwd = 2)
}




# -- (2b) Mark the true parameter on the floor


points3d(
  x = par_true[1],
  y = par_true[2],
  z = z_floor,
  color = "red",
  size = 10
)

points3d(
  x =hat.theta12[1], 
  y = hat.theta12[2], 
  z = z_floor,
  size = 10, 
  col = "blue", 
  cex = 1
)

```


```{r}
library(htmlwidgets)  # for saveWidget

# Convert to a widget
my_widget <- rglwidget(width = 800, height = 800)

# Save the widget to an HTML file
saveWidget(my_widget, file = "my3Dplot.html")

```



#Confidence Interval

#sample variance----hessian function
```{r}
#negative hessian
#hessian function
hessian<-function(sum.ei,Mt,par1,par2,S0){
  
  the1<- par1
  the2<- par2
  
  x0<-S0[1]
  s.ei<-sum.ei
  
  
 dev_theta_1.square<- -2*the2*x0/(the1^3)+ 2*the2*x0*the1^{-3}*exp(-the1*Mt) + 2*Mt*the2*x0*the1^{-2}*(exp(-the1*Mt))+ Mt^2*the2*x0*exp(-the1*Mt)/the1
 
  dev_the2.square<- -sum.ei*(the2^{-2})
  
  dev_theta1_theta2<- -the1^{-2}*(exp(-the1*Mt)-1)-Mt*the1^{-1}*exp(-the1*Mt)
 
  
 local.hessian<-dev_theta_1.square*dev_the2.square-dev_theta1_theta2^2
  
 H<-matrix(c(dev_theta_1.square, dev_theta1_theta2, dev_theta1_theta2, dev_the2.square),nrow = 2, byrow = TRUE)
 
  sigma<-solve(H)
  
return(  sigma)
  
}


```




#Asymptotic Confidence Interval


$$
\hat{\boldsymbol{\theta}_i}\pm \sqrt{\chi^2(\alpha,2)*C_{i,i}}\\
C_{i,i}=2*H^{-1}_{ii}
$$


```{r}
the1<-hat.theta12[1]
the2<-hat.theta12[2]

x0=S0[1]/N

sigma<-hessian(sum.ei=sum.ei,Mt=Mt,par1=the1,par2=the2,S0=S0/N)

c((hat.theta12  -sqrt(qchisq(0.95, df = 1)*(-1)*eigen(Sigma)$values*2))[1],(hat.theta12 +sqrt(qchisq(0.95, df = 1)*(-1)*eigen(Sigma)$values*2))[1])
#hat_theta_1

c((hat.theta12  -sqrt(qchisq(0.95, df = 1)*(-1)*eigen(Sigma)$values*2))[2],(hat.theta12 +sqrt(qchisq(0.95, df = 1)*(-1)*eigen(Sigma)$values*2))[2])

#hat_theta_2

```





```{r}


pdf("myContour-with-ACI.pdf", width = 8, height = 6)
# Zoom range around (0.11, 1)
x_zoom <- seq(0.02, 0.2, length.out = 200)
y_zoom <- seq(0.9,1.05, length.out = 200)


# Recompute z for the zoomed grid
z_zoom <- outer(x_zoom, y_zoom, loglikelihood.full.data)



contour(
  x_zoom, 
  y_zoom, 
  z_zoom,
  xlab = expression(theta[1]),  # θ₁
  ylab = expression(theta[2]),  # θ₂
  main = "Log-Likelihood Contour Plot with Asymptotic Confidence Interval",
  nlevels = 100
)

# Mark the true parameter in 2D
points(
  x = par_true[1], 
  y = par_true[2], 
  pch = 19, 
  col = "red", 
  cex = 1
)

points(
  x = hat.theta12[1], 
  y = hat.theta12[2], 
  pch = 19, 
  col = "blue", 
  cex = 1
)

text(
  x = par_true[1],
  y = par_true[2], 
  labels = bquote(
    theta^{"true"} == "(" ~ .(round(par_true[1], 3)) ~ "," ~ .(round(par_true[2], 3)) ~ ")"
  ),
  pos = 4,
  col = "red"
)

text(
  x = hat.theta12[1],
  y = hat.theta12[2], 
  labels = bquote(
    theta^{"mle"} == "(" ~ .(round(hat.theta12[1], 3)) ~ "," ~ .(round(hat.theta12[2], 3)) ~ ")"
  ),
  pos = 4,
  col = "blue"
)

# Example covariance matrix
Sigma <- (-1)*sigma

# Generate and overlay the confidence region
ellipse_points <- function(center, covmat, level = 0.95, npoints = 200) {
  r <- sqrt(qchisq(level, df = 2))
  eig <- eigen(covmat)
  evec <- eig$vectors
  evals <- eig$values
  axes <- r * sqrt(evals)
  theta_seq <- seq(0, 2*pi, length.out = npoints)
  circle <- cbind(cos(theta_seq), sin(theta_seq))
  ellipse <- t(evec %*% (diag(axes) %*% t(circle)))
  ellipse[,1] <- ellipse[,1] + center[1]
  ellipse[,2] <- ellipse[,2] + center[2]
  ellipse
}

# Add the confidence region
confidence_region <- ellipse_points(hat.theta12, Sigma, level = 0.95)
polygon(confidence_region, col = rgb(1, 0.5, 0.2, 0.3), border = NA)
```





```{r}


# Zoom range around (0.11, 1)
#x_zoom <- seq(par_true[1] - 0.02, par_true[1] + 0.02, length.out = 200)
#y_zoom <- seq(par_true[2] - 0.1, par_true[2] + 0.1, length.out = 200)

pdf("myContour-with-ACI.pdf", width = 8, height = 6)
# Zoom range around (0.11, 1)
x <- seq(theta1_lower- 0.02, theta1_upper + 0.02, length.out = 200)
y <- seq(theta2_lower - 0.1, theta2_upper+ 0.1, length.out = 200)


# Recompute z for the zoomed grid
z<- outer(x, y, loglikelihood.full.data)

contour(x, y, z, nlevels = 100,
        xlab = expression(theta[1]),
        ylab = expression(theta[2]),
        main = "Log-Likelihood Contour Plot")

# Mark MLE
points(hat.theta12[1], hat.theta12[2], pch = 19, col = "blue")
points( par_true[1],  par_true[2], pch = 19, col = "red",  cex = 1)

text(
  x     = hat.theta12[1],
  y     = hat.theta12[2],
  labels = bquote(
  theta^{"mle"} == "(" ~ .(round(hat.theta12[1], 3)) ~ "," ~ .(round(hat.theta12[2], 3)) ~ ")"
),
  pos   = 4,
  col   = "blue"
)
text(
  x     = par_true[1],
   y = par_true[2], 
  labels = bquote(
  theta^{"true"} == "(" ~ .(round(par_true[1], 3)) ~ "," ~ .(round(par_true[2], 3)) ~ ")"
),
  pos   = 3,
  col   = "red"
)
# Remove or comment out the abline(...) lines
# Then draw the rectangle:

rect(
  xleft   = theta1_lower,
  ybottom = theta2_lower,
  xright  = theta1_upper,
  ytop    = theta2_upper,
  border  = "blue",
  lty     = 2,
  lwd     = 2
)
```







#The confidence Interval--likelihood-based Confidence Intervals

$$

\{ (\theta_1,\theta_2):\, 2|l(\hat{\theta})-l(\theta_1,\theta_2)| \leq \chi_{2,1-\alpha^2 }\}

$$
The sample derivance follows the $\chi^2(1,1-\alpha)$ distribution for one of the parameter. We set the 95/% of contour boundary as the $\chi_{2,1-\alpha^2}^2 $ plane.

```{r}
# Maximum log-likelihood value:
llhat <- loglikelihood.full.data(hat.theta12[1], hat.theta12[2])

z <- outer(x_zoom, y_zoom, FUN = Vectorize(loglikelihood.full.data))

dev <- 2 * (llhat - z)

crit_val <- qchisq(0.95, df = 2)  # ~ 5.9915 for 95% CI

```

```{r}
pdf("myContour-with-chi-squre-CI.pdf", width = 8, height = 6)


#Basic 2D contour of the log-likelihood
contour(
  x_zoom,y_zoom, z,
  xlab  = expression(theta[1]),    # θ₁
      ylab = expression(theta[2]),    # θ₂
  main = "Log-Likelihood Contours + 95% CI",
  nlevels = 80
)

# Add the boundary of the region: dev = crit_val
contour(
 x_zoom, y_zoom, dev,
  levels = crit_val,       # the contour level
  drawlabels = FALSE, 
  add = TRUE,              # add to existing plot
  col = "red", 
  lwd = 2
)
points(
  x = par_true[1], 
  y = par_true[2], 
  pch = 19, 
  col = "red", 
  cex = 1
)

points(hat.theta12[1], hat.theta12[2], 
       pch = 19, col = "blue", cex = 0.5)



text(
  x     = hat.theta12[1],
  y     = hat.theta12[2],
  labels = bquote(
  theta^{"mle"} == "(" ~ .(round(hat.theta12[1], 3)) ~ "," ~ .(round(hat.theta12[2], 3)) ~ ")"
),
  pos   = 4,
  col   = "blue"
)
text(
  x     = par_true[1],
   y = par_true[2], 
  labels = bquote(
  theta^{"true"} == "(" ~ .(round(par_true[1], 3)) ~ "," ~ .(round(par_true[2], 3)) ~ ")"
),
  pos   = 4,
  col   = "red"
)

#Optional with the shade


# Get contour lines for dev = crit_val
region_line <- contourLines(x_zoom,y_zoom, dev, levels = crit_val)

# region_line is a list of line(s). Typically, for a single closed contour,
# it's region_line[[1]]. Let's assume there's only one closed boundary:

cl <- region_line[[1]]
x_poly <- cl$x
y_poly <- cl$y

# Now fill that region. 
# We fill the region *outside* the contour or inside depending on your data shape.
# Usually, we fill inside by coloring it and rely on plotting order to see it.
polygon(x_poly, y_poly, col = rgb(1, 0, 0, alpha=0.2), border = NA)
```

